A fast and Low-Overhead Technique to Secure Programs Against Integer Overflows

### 2012-11-09 doc/Manuscript/CGO/paper.tex
Put the paper in the CGO format.

\documentclass[preprint]{sigplanconf}[10pt]

@@

\begin{abstract}
The integer primitive type has upper and lower bounds in many programming
languages, including C, and Java.
These limits might lead programs that manipulate large integer numbers to
produce unexpected results due to overflows.
There exists a plethora of works that instrument programs to track the occurrence
of these overflows.
In this paper we present the first algorithm that uses static range analysis to
avoid this instrumentation whenever possible.
Our range analysis contains novel techniques, such as a live range splitting
strategy that refines the precision of our results.
We have used this algorithm to avoid some checks created by a dynamic instrumentation
library that we have implemented in LLVM.
This framework has been used to detect overflows in hundreds of C/C++ programs.
As a testimony of its effectiveness, our range analysis has been able to avoid
25\% of all the overflow checks necessary to secure the C programs in the
LLVM test suite.
This reduction has reduced 50\% of the runtime overhead of instrumentation.
\end{abstract}

@@

\subsection{The Danger that Integer Overflows Represent}
\label{sub:danger}

% A brief overview about integer overflows.
In addition to being a source of several bugs that are hard to identify, integer 
overflows can represent a serious security risk.
As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
sources of vulnerabilities in widely used software, such as OpenSSH and
Firefox.
These vulnerabilities could give an adversary the opportunity to execute
arbitrary code in the compromised application. 
We will demonstrate this with an example.
The $n$-bits two's complement arithmetics wraps around at multiples of
$2^n$.
As an example the representation of $16 \times 16$ in 8-bits two's complement
is $0$, because $0 \equiv (16\times16) \mod 2^8$.
Programmers can use this well-defined behavior in benign ways.
For instance, the wrapping arithmetics provides a cheap $\mod 2^n$ operator,
which is used in the implementation of hash-functions or random number
generators.

@@

% An example of vulnerability.
Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
overflow vulnerability.
This vulnerability allows an adversary to perform a buffer overrun attack
on a program that is apparently guarded against this type of exploit.
For simplicity, we will use the C datatype \texttt{char}, which represents
8-bit long numbers.
Examples of actual vulnerabilities, with larger primitive types, are given
by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
Zhang {\em et al.}~\cite{Zhang10}.
The function \texttt{read\_matrix} reads a matrix of bytes from a file.
This matrix is stored as a linear sequence of bytes in a vector \texttt{out}.
The first and second bytes in the file, e.g., \texttt{w} and \texttt{h},
determine the number of rows and columns in the matrix.
If the \texttt{char} datatype could represent arbitrarily long numbers, than this
function would be safe against buffer overruns, because it would only write data
on allocated memory.
However, if, for instance, we have that \texttt{w}$= 16$ and \texttt{h}$= 16$,
then \texttt{s}$= 0$, as pointed before.
In this case, a buffer of length zero would be allocated, and all the data
stored in the file would overwrite memory in the stack of activation records.
By carefully crafting an input file, an adversary can, in this way, overwrite
the return address of \texttt{read\_matrix}, diverting program's execution to
one of \texttt{libc}'s function, such as \texttt{telnet}, for instance.

@@

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set \varset{C} of constraints ranging over a set
\varset{V} of variables. \\
\textbf{Output:} a mapping I such that, for any variable
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

@@

We find the ranges of the variables in each strongly connected component in
three phases.
First we determine the growth pattern of each variable in the component via
widening.
In the second step, once we know in which directions the values stored in each
variable might grow, we replace futures by actual bounds.
Finally, a narrowing phase starting from conditional tests improves the
precision of our results.

@@

>>>>

\documentclass{sigplanconf}[10pt]

@@

\begin{abstract}
The integer primitive type has upper and lower bounds in many programming
languages, including C, and Java.
These limits might lead programs that manipulate large integer numbers to
produce unexpected results due to overflows.
There exists a plethora of works that instrument programs to track the occurrence
of these overflows.
In this paper we present an algorithm that uses static range analysis to
avoid this instrumentation whenever possible.
Our range analysis contains novel techniques, such as a notion of ``future"
bounds to handle comparisons between variables.
We have used this algorithm to avoid some checks created by a dynamic instrumentation
library that we have implemented in LLVM.
This framework has been used to detect overflows in hundreds of C/C++ programs.
As a testimony of its effectiveness, our range analysis has been able to avoid
25\% of all the overflow checks necessary to secure the C programs in the
LLVM test suite.
This reduction has reduced 50\% of the runtime overhead of instrumentation.
\end{abstract}

@@

\noindent
\textbf{The Danger that Integer Overflows Represent: }
Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
overflow vulnerability.
This vulnerability allows an adversary to perform a buffer overrun attack
on a program that is apparently guarded against this type of exploit.
We will use the C datatype \texttt{char}, which represents
signed 8-bit numbers, so that the interested reader can easily
reconstruct this exploit.
Examples of actual vulnerabilities, with larger primitive types, are given
by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12},
and Zhang {\em et al.}~\cite{Zhang10}.
The function \texttt{read\_matrix} copies the linearized matrix \texttt{data}
into \texttt{buf}, to later process it, in line 12.
The arguments \texttt{w} and \texttt{h} determine the number of rows and
columns in the matrix.
If the \texttt{char} datatype could represent arbitrarily long numbers, than
this function would be safe against buffer overruns, due to the test at
line 3.
However, if, for instance, we have that \texttt{w}$= 6$ and \texttt{h}$= 22$,
then \texttt{buf\_size}$= -124$, due to the wrapping semantics of
integer arithmetics in C.
In this case, twelve words in the activation record of \texttt{read\_matrix}
would be overwritten.
By carefully crafting the contents of \texttt{data}, an adversary can, in this
way, change the return address of \texttt{read\_matrix}, diverting program's
execution to a system's function, such as \texttt{telnet}, for instance.

@@

@@

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set \varset{C} of constraints ranging over a set
\varset{V} of variables. \\
\textbf{Output:} a mapping I such that, for any
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

@@

We find the ranges of the variables in each strongly connected component in
three phases.
First we determine the growth pattern of each variable in the component via
widening.
In the second step, we replace future bounds by actual limits.
Finally, a narrowing phase starting from conditional tests improves the
precision of our results.

@@

\acks
We thank Douglas do Couto for helping us with the LLVM code base.
We also thank the anonymous reviewers for their useful comments and
suggestions.
This project is supported by FAPEMIG, grant 16899*01.
